---
title: "HW10 Q13-14"
author: 'Sam Altshuler (PID: A59010373)'
date: "2/21/2022"
output: pdf_document
---

# Section 4: Population Scale Analysis 
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.
This is the final file you got ( https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt ). The first column is sample name, the second column is genotype and the third column are the expression values. Open a new RMarkdown document in RStudio to answer the following two questions. Submit your resulting PDF report with your working code, output and narrative text answering Q13 and Q14 to GradeScope.

* Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 
  + The sample size for the A/A genotype is n = 108, the sample size for G/A is n = 233, and for G/G is n = 121. The median expression levels for each genotype is 31.248, 25.064, and 20.074 (A/A, A/G, and G/G respectively).
```{r}
library(dplyr)

gen_dat <- read.table(url("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"))

gen_box <- boxplot(gen_dat$exp ~ gen_dat$geno)
gen_box
```
* Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?
  + From the boxplot below, it appears that the SNP does negatively affect the expression of ORMDL3. The G/G expression data is visually much lower than its A/A counterpart. This can also be seen in the answer above due to the differences in median expression levels.
```{r}
library(ggplot2)

ggplot(gen_dat, aes(x= geno, y = exp)) +
  geom_boxplot()+
  labs(title = "Expression levels for ORMDL3 SNP")+
  xlab("Genotype") +
  ylab("Expression Level")
```

