---
title: "Machine Learning Lab 7"
author: 'Sam Altshuler (PID: A59010373)'
date: "2/10/2022"
output: pdf_document
---
# PCA of UK food data

## Import the UK foods dataset
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
# How many rows and columns does it have?
dim(x)
```
The dataset has 17 rows and 5 columns. However the first column is the rownames.
## Check the data
```{r}
head(x)
```

Set the first column as rownames
```{r}
rownames(x) <- x[,1]
#remove the first row since it was set to the rownames
x <- x[,-1]
head(x)
```
Now see what `dim()` returns
```{r}
dim(x)
```
You could also rename the rownames in the initial read.csv() by specifying which columns are the names.
```{r}
x <- read.csv(url, row.names = 1)
head(x)
```
This way is prefered so that you don't run the risk of deleting any of the data when doing the first style of data transformation.

## Spotting major differences and trends

Barplot of the data, it's hard to get any valuable information between different components.
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
Change the barplot to have them stacked for each country.

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```
Pairwise plots might be helpful
```{r}
pairs(x, col=rainbow(10), pch=16)
```
This plot compares between the two countries the amount of consumption per category. If a value is on the diagonal line, it means they have the same amount of consumption for that food category. This is a fold-change of zero. 
However, it's still hard to determine main differences between one country from the other countries.

## PCA to the rescue

Do PCA of this 17 dimension UK food data. The main function in base R is called `prcomp()`
```{r}
# Need to transpose x to make it in the correct format for prcomp()
pca <- prcomp(t(x))
summary(pca)

```
The `prcomp()` function returns a list of objects
```{r}
plot(pca)
```
The "PCA plot" is also known as a pca score plot. It is a plot of PC1 v PC2. Basically a new PCA axis to view the data.

```{r}

attributes(pca)
```
Pay attention to the "x" attribute which is a matrix of the data (`pca$x`).
```{r}
pca$x

plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2") + text(pca$x[,1], pca$x[,2], labels = colnames(x), col =c("gold", "red", "blue", "dark green"))
```

## Digging Deeper

```{r}
# focusing on PC1 since it accounts for >90% of all variance in the dataset
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```
Now for PC2
```{r}
# focusing on PC1 since it accounts for >90% of all variance in the dataset
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```

Biplots are another way to view this data. The arrows from the center show the amount of variance that single dimension is responsible for in each principal component.
```{r}
biplot(pca)

```

#PCA of RNA-seq data

Import the expression data.
```{r}

url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
dim(rna.data)
```
 There are 100 genes and 10 samples in this data set (genes are rows and samples are columns).
 
Perform the PCA analysis and do a simple plot.
```{r}
pca <- prcomp(t(rna.data), scale = TRUE)
summary(pca)
#simple plot
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2")
```
About 93% of the variance can be shown in the PC1 dimension.
Plot the Scree-plot to show the elbow point.
```{r}
plot(pca, main="Quick scree plot")
```
We can also generate our own Scree plot.
```{r}
# Variance caught per principal component
pca.var <- pca$sdev^2

# percent variance is usually easier to look at than just variance
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```
Now use this data to make our own scree plot. 
```{r}
barplot(pca.var.per, main = "Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab = "Principal Component", 
        ylab = "Percent Variance")
```

Now we can make the PCA plot a bit easier to look at. 

```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```
Use ggplot because it's more fun.

```{r}
# Load in the ggplot package
library(ggplot2)
df <- as.data.frame(pca$x)

# Basic plot once again!
ggplot(df) +
  aes(PC1, PC2) +
  geom_point()
```
Add in some colors and other aesthetics.

```{r}
# Add a wt and ko condition column
df$samples <- colnames(rna.data)
df$condition <- substr(colnames(rna.data), 1, 2)

#Plot with labels and colors
p <- ggplot(df) +
  aes(PC1, PC2, label = samples, col = condition) +
  geom_label(show.legend = FALSE)
p
```
Add some more labels
```{r}
p + labs(title = "PCA of RNASeq Data",
         subtitle = "PC1 separates wild-type from knock out conditions",
         x = paste0("PC1 (", pca.var.per[1], "%)"),
         y = paste0("PC2 (", pca.var.per[2], "%)"),
         caption = "BIMM143 example data") +
  theme_bw()
```

