---
title: "08_mini_project"
author: 'Sam Altshuler (PID: A59010373)'
date: "2/11/2022"
output: pdf_document
---
# Exploratory Data Analysis
## Preparing the data
```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
head(wisc.df)
```
Remove the diagnosis column and save as a new df to not overwrite original data.
```{r}
wisc.data <- wisc.df[,-1]
# Diagnosis vector with original data diagnosis
diagnosis <- wisc.df$diagnosis
```

# Exploratory data analysis


* Q1. How many observations are in this dataset?
```{r}
dim(wisc.data)
```
  + There are 569 separate observations 
  
* Q2. How many of the observations have a malignant diagnosis?
```{r}
length(diagnosis[diagnosis == "M"])
```
  + There are 212 observations with a malignant diagnosis 
* Q3. How many variables/features in the data are suffixed with _mean?
```{r}
length(grep("_mean", colnames(wisc.data)))
```
  + There are 10 variables that are suffixed with "_mean".

# Principal Component Analysis
Next step in the analysis is to perform a PCA on `wisc.data`. 
Data needs to be rescaled if:
* The input variables use different units of measurement
* The input variables have significantly different variances

```{r}
# Check column means and standards deviations
colMeans(wisc.data)
# apply standard deviation function to every column in the dataframe
apply(wisc.data, 2, sd)
```
Based on the means and standard deviations, it appears that the dataset has very different variances and likely different units of measurement (for example smoothness and symmetry are likely not in the same units). 
PCA with scaling:
```{r}
wisc.pr <- prcomp(wisc.data, scale. = TRUE)
# summary of the results
summary(wisc.pr)
```


* Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
  + PC1 captures 44.27% of the original variance.
* Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
  + You need to include the first three PCs (1-3) in order to describe at least 70% of the original variance. 
* Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
  + At least 7 principal components are needed to describe >90% of the original variance.
  
## Interpreting PCA results 

Visualize the PCA using a biplot
```{r}

biplot(wisc.pr)
```

* Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
  + This plot is very hard to read, you cannot even separate out the observations into readable labels. There is just too much crowding. 
  
Let's create a standard scatter plot using PC1 as the x and PC2 as the y axis.

```{r}
plot(wisc.pr$x, 
     col = as.factor(diagnosis), 
     xlab = "PC1", 
     ylab = "PC2")
```
* Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
```{r}
plot(wisc.pr$x[,c(1,3)], 
     col = as.factor(diagnosis), 
     xlab = "PC1", 
     ylab = "PC3")
```
 + These plots seem relatively similar, however that is likely because PC1 explains much of the variance. However, the plot comparing PC1 v PC2 has a more clean line between the benign and malignant cells since together they explain 63% of the total variance while PC1 and PC3 only explain 55%. This is seen by the overlap in colors around PC1 = 0 in the PC1 v PC3 graph.
 
 
 Use ggplot2 to create some better looking figures.
```{r}
# load in the package
library(ggplot2)
#create a dataframe for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Scatter plot colored by diagnosis
ggplot(df) +
  aes(PC1, PC2, col = diagnosis)+
  geom_point()

```
 
 
## Variance Explained

Produce the scree plots to show proportion of variance explained as number of PCs increases.

```{r}
# Calculate the variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

 Calculate percent variance to understand the amount of variance that can be explained by each PC.
```{r}
pve <- pr.var/sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

Or this can be shown as a bargraph:
```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```
There isn't an easy elbow to see but I would say the elbow is around PC5. 

## Communicating PCA results


* Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?
  + The loading vector shows the covariance between the original value and the scaled principal component. For example, the concave.points_mean loading vector is -0.2608, which means this is the value that is multiplied by the concave.points_mean to determine it's contribution to PC1.

* Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?
  + There is a minimum 5 PCs needed to explain at least 80% of the variance of the data. 


# Prediction

Add in new data
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
Plot these predictions on the original PCA data.
```{r}
plot(wisc.pr$x[,1:2], col=as.factor(diagnosis))
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
 
* Q18. Which of these new patients should we prioritize for follow up based on your results?
  + You should prioritize patient 2 because their data point appears to be within the malignant clusters.

 