---
title: "Class06 Lab Functions"
author: 'Sam Altshuler (PID: A59010373)'
date: "2/4/2022"
output: pdf_document
---

In this class we are going to learn all about functions in R. 


First we will write a function to grade some student scores.
Use ctl+alt+I to insert code block.

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

```

Calculate the scores of the student assignments for each student. Usually, we could just use a mean() function, but you cannot take a mean of a vector that contains NA values.

```{r}
# Results in NA values
mean(student1)
mean(student2)
mean(student3)

# Strip NA values using na.rm = TRUE
mean(student1, na.rm =T)
mean(student2, na.rm =T)
mean(student3, na.rm =T)
```
mean(x, na.rm = T/F) will either remove or keep all NA values for a vector X.
This isn't fair for student 3 since they only did one assignment and will still get a 90 for their overall homework grade. This could be fixed by setting NA to zero.

To find NA values (not available, AKA missing values), use `is.na()` function. 

```{r}
student2
# We know that only NA value for student 2 is in position 2
is.na(student2)
```
Quick review of logical vectors
```{r}
x <- 1:5
x < 2
x == 3
```

Set all NA values to zero
```{r}
# use a temp variable to not override the original data
n_student <- student3
n_student[is.na(student3)] <- 0
student3
n_student

```
We can use min() to find smallest value

```{r}
# Return the smallest value
min(student1)
# Return index (position) of the smallest value
which.min(student1)
```

Put it all together into a function! This function should drop the single lowest score per student and then determine the overall score.

```{r}

grade <- function(x){
  # Replace all NA values with zero
  x[is.na(x)] <- 0
  # Remove the smallest value
  y <- x[-which.min(x)]
  # Average out the score
  avg <- mean(y)
  # Return the average
  
  return(avg)
}
```
Test out the function.

```{r}
grade(student1)
grade(student2)
grade(student3)
```
## Grade the Class

Input the dataset (it must be in the same workspace/directory) or use `read.csv()` with the file url as a string.

```{r}
# use read.csv to insert the gradebook as a dataframe
gradebook <- read.csv("https://tinyurl.com/gradeinput", row.names = 1)
gradebook
```
> Q2. Use the grade() function to determine who is the top scoring student

```{r}
score <- apply(gradebook, 1, grade)
score
which.max(score)
```
Student 18 is the top scoring student according to this grade book with a score of 94.50

> Q3. Which homework is the hardest for the students (which homework has the lowest overall score)?

```{r}

# apply the mean function to every column
# use na.rm because NA values are outliers  
# and should just be ignored
hw <- apply(gradebook, 2, mean, na.rm = TRUE)
hw
which.min(hw)
```
Homework 3 has the lowest mean score if NA's are removed. But... homework 2 has the lowest overall score as shown by the median in the boxplot below. 

```{r}
boxplot(gradebook)
```
Calculate the median to determine the homework that was the toughest as a better way to determine toughness than the mean (since outliers can affect the mean as shown by the boxplot above).
```{r}
hw_new <- apply(gradebook, 2, median, na.rm = TRUE)
hw_new
which.min(hw_new)
```
So homework 2 was the toughest with a median score of 72.5


